Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-02-20 paper-LIN-PQ.tex
change paper-LIN-PQ to seperate sections into individual files(just a start)DDDDD

\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{color}
%\usepackage{txfonts}
%\usepackage{pxfonts}
%\usepackage{mathabx}
%\usepackage{makeidx}  % allows for indexgeneration
\usepackage{verbatim}
\usepackage{url}
%\usepackage{hyperref}
\usepackage{times}
\usepackage{ulem}    %(\emph{....} in the texfile) is underlined instead of italic
\normalem
\usepackage{enumerate}
%\usepackage[square, comma, sort&compress, numbers]{natbib}

@@

\author {Chao Wang \and Ahmed Bouajjani \and Constantin Enea}
\institute{Institut de Recherche en Informatique Fondamentale, \\Univ. Paris Diderot (Paris 7)
}

@@

\begin{abstract}
Abstract
\end{abstract}

\forget{
\noindent Keywords: weak memory model, $\textit{linearizability}$,
$\textit{TSO-to-TSO linearizability}$
}

\section{Introduction}
\label{sec:introduction}
Introduction

\smallskip

\noindent {\bf Related work}
%\label{sec:related work}
\noindent Related work

\section{Inductive Rules for Priority Queues}
\label{sec:inductive rules for priority queues}

given a sequence $l$ of priority queue operations, we say that $l$ is data-differentiated, if for each item $d \in \mathcal{D}$, $\textit{put}(d,\_)$ appears at most once in $l$. Note that a data-differentiated sequence $l$ may contains more than one items with a same priority. The subset of data-differentiated operation sequences of a set $S$ is denoted by $S_{\neq}$. For priority queue, the renaming functions rename only the items and keep the priorities unchanged. For priority queue, we modify the notion of data-independence \cite{Wolper:1986} and use data-differentiated instead of differentiated:

%\vspace{-6pt}
\begin{definition}[data-independence of priority queue]\label{def:priority-value data-independence}
The set $S$ of sequences of priority queue operations (resp., histories)  is data independent, if:
\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] for all $l \in S$, there exists $l' \in S_{\neq}$, and a renaming $r$, such that $l=r(l')$,

\item[-] for all $l \in S$ and for all renaming $r$, $r(l) \in S$.
\end{itemize}
\end{definition}

Similar as in \cite{Bouajjani:2015}, we consider only completed executions, where each call action has a corresponding return action. This simplification is sound when implementation methods can always make progress in isolation.

For the inductive rules of priority queue, our rules need to concern the cost of operations. Formally, the inductive rules for priority queues is as follows:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $R_{\textit{pr0}} \equiv \epsilon \in \textit{PQueue}$

\item[-] $R_{\textit{pr1}} \equiv (u \cdot v \cdot w \in \textit{PQueue}) \wedge (u \cdot v \cdot w \in (\textit{put},\textit{rm})^*) \wedge (\textit{pri} >$ priorities of $u \cdot v \cdot w ) \wedge (\textit{matched}(u \cdot v, \textit{put}, \textit{rm}) ) \Rightarrow (u \cdot \textit{put}(\textit{itm},\textit{pri}) \cdot v \cdot \textit{rm}(\textit{itm}) \cdot w \in \textit{PQueue})$

\item[-] $R_{\textit{pr2}} \equiv (u \cdot v \cdot w \in \textit{PQueue}) \wedge (u \cdot v \cdot w \in (\textit{put},\textit{rm})^*) \wedge (\textit{pri} =$ the maximal priority of $u \cdot v \cdot w ) \wedge (\textit{pri} >$ priorities of unmatched items in $u \cdot v \cdot w) \wedge (\textit{matched}(u \cdot v, \textit{put}, \textit{rm}) ) \wedge ($all $\textit{put}$ operation with priority $\textit{pri}$ is in $u) \Rightarrow (u \cdot \textit{put}(\textit{itm},\textit{pri}) \cdot v \cdot \textit{rm}(\textit{itm}) \cdot w \in \textit{PQueue})$

\item[-] $R_{\textit{pr3}} \equiv (u \cdot v \in \textit{PQueue}) \wedge ( u \cdot v \in (\textit{put},\textit{rm})^*) \wedge (\textit{pri} =$ the maximal priority of items in $u \cdot v) \wedge (\textit{pri} >$ priorities of matched items in $u \cdot v \cdot w ) \Rightarrow (u \cdot \textit{put}(\textit{itm},\textit{pri}) \cdot v \in \textit{PQueue})$

\item[-] $R_{\textit{pr4}} \equiv (u \cdot v \in \textit{PQueue}) \wedge ( u \cdot v \in (\textit{put},\textit{rm})^*) \wedge (\textit{pri} =$ the maximal priority of items in $u \cdot v) \wedge (\textit{pri} =$ the maximal priority of matched items in $u \cdot v \cdot w ) \wedge ($all matched $\textit{put}$ with priority $\textit{pri}$ is in $u) \wedge ($all unmatched $\textit{put}$ with priority $\textit{pri}$ is in $u) \Rightarrow (u \cdot \textit{put}(\textit{itm},\textit{pri}) \cdot v \in \textit{PQueue})$

\item[-] $R_{\textit{pr5}} \equiv (u \cdot v \in \textit{PQueue}) \wedge (\textit{matched}(u,\textit{put},\textit{rm}) ) \Rightarrow (u \cdot \textit{rmEmpty} \cdot v \in \textit{PQueue})$
\end{itemize}

Given a sequence $l$ of possibly $\textit{put}$, $\textit{rm}$ and $\textit{rmEmpty}$ operations, we say that $l$ contains unmatched $\textit{rm}$ operations, if the number of $\textit{rm}(d)$ operations is larger than the number of $\textit{put}(d,\_)$ operations for some $d$ in $l$. By data-independence, we can detect the existence of unmatched $\textit{rm}$ operations by automata. A automaton $\mathcal{A}_{\textit{unmRm1}}$ is given in \figurename~\ref{fig:automata for unmatched rm1}, where $c_1=\textit{cal}(\textit{put},a,\textit{anyPri}),\textit{cal}(rm),\textit{ret}(\textit{put}),\textit{ret}(\textit{rm},a)$, $c_2=c_1 + \textit{ret}(\textit{rm},b)$, and $c_3 = c_1 + \textit{cal}(\textit{put},b,\textit{anyPri}) + \textit{ret}(\textit{rm},b)$. Here $\textit{anyPri}$ is a specific predicates that is satisfied by any priority. $\mathcal{A}_{\textit{unmRm1}}$ is used to detect the existence of a item that has been removed but has never been inserted.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5 \textwidth]{PIC_AUTO_UNMATCHED_RM1.pdf}
%\vspace{-10pt}
  \caption{Automaton $\mathcal{A}_{\textit{unmRm1}}$}
  \label{fig:automata for unmatched rm1}
\end{figure}

An automaton $\mathcal{A}_{\textit{unmRm2}}$ is given in \figurename~\ref{fig:automata for unmatched rm2}, where $c_1 = \textit{cal}(\textit{put},a,\textit{anyPri}),\textit{cal}(rm),$ $\textit{ret}(\textit{put}),\textit{ret}(\textit{rm},a)$, and $c_1 = c_1 + \textit{ret}(\textit{rm},b)$.  $\mathcal{A}_{\textit{unmRm1}}$ is used to detect the existence of a item that has been removed more than once but has been inserted once.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5 \textwidth]{PIC_AUTO_UNMATCHED_RM2.pdf}
%\vspace{-10pt}
  \caption{Automaton $\mathcal{A}_{\textit{unmRm2}}$}
  \label{fig:automata for unmatched rm2}
\end{figure}

The following lemma states that we can detect the existence of unmatched $\textit{rm}$ by $\mathcal{A}_{\textit{unmRm1}}$ and $\mathcal{A}_{\textit{unmRm2}}$.

\begin{restatable}{lemma}{DetectUnmatchedRmByAutomata}
\label{lemma:detect unmatched rm by automata}

Given a data-independent $\mathcal{I}$, there is some unmatched $\textit{rm}$ in some sequence of $\mathcal{I}$, if and only if $( \mathcal{I} \cap \mathcal{A}_{\textit{unmRm1}} \neq \emptyset) \vee ( \mathcal{I} \cap \mathcal{A}_{\textit{unmRm2}} \neq \emptyset )$.
\end{restatable}

\begin {proof}

For the $\textit{if}$ direction, assume that $( \mathcal{I} \cap \mathcal{A}_{\textit{unmRm1}} \neq \emptyset) \vee ( \mathcal{I} \cap \mathcal{A}_{\textit{unmRm2}} \neq \emptyset )$. Then, there exists a sequence $l \in \mathcal{I}$ and $l$ is accepted by $\mathcal{A}_{\textit{unmRm1}}$ or $\mathcal{A}_{\textit{unmRm2}}$. If $l$ is accepted by $\mathcal{A}_{\textit{unmRm1}}$, then $l$ contains at least one $\textit{rm}(b)$ operation and no $\textit{put}(b,\_)$ operation; Else, if $l$ is accepted by $\mathcal{A}_{\textit{unmRm2}}$, then $l$ contains at least two $\textit{rm}(b)$ operations and one $\textit{put}(b,\_)$ operation. In both cases, $l$ contains unmatched $\textit{rm}$ operations.

To prove the $\textit{only if}$ direction assume that there exists a sequence $l \in \mathcal{I}$ that contains unmatched $\textit{rm}$ operations. By data-independence, there exists data-differentiated sequence $l'$ and renaming function $r$, such that $l=r(l')$. We need to prove that in $l'$, one of the following cases holds: (1) $l'$ contains $\textit{rm}(d_1)$ for some $d_1$ and does not contain any $\textit{put}(d_1,\_)$, and (2) $l'$ contains one $\textit{put}(d_1,\_)$ operation and more than one $\textit{rm}(d_1)$ operations. We prove this by contradiction. Assume by contradiction that both cases do not hold. Then, for each value $d$ of $l'$, there is either one $\textit{put}(d,\_)$ and one $\textit{rm}(d)$ in $l'$, or one $\textit{put}(d,\_)$ and no $\textit{rm}(d)$ in $l'$. It is obvious that such $l'$ does not contain unmatched $\textit{rm}$ operation, and renaming a sequence that does not contain unmatched $\textit{rm}$ we can only obtain another sequence that does not contain unmatched $\textit{rm}$. This contradicts that $l=r(l')$ has unmatched $\textit{rm}$.

Let $l''$ be obtained from $ll$ by renaming $d_1$ to $b$ and renaming other value to $a$. If the first case holds, $l''$ is accepted by $\mathcal{A}_{\textit{unmRm1}}$; Otherwise, the second case holds, and $l''$ is accepted by $\mathcal{A}_{\textit{unmRm2}}$. \qed
\end {proof}

Therefore, from now on, we can safely assume that there are no unmatched $\textit{rm}$ operations. The notion of $\textit{last}$ is defined as follows:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $l$ contains $\textit{rmEmpty}$ operations, then $\textit{last}(l) = R_{\textit{pr5}}$.
\item[-] Else, $l$ does not contain any $\textit{rmEmpty}$ operation.
    \begin{itemize}
    \setlength{\itemsep}{0.5pt}
    \item[-] If the maximal priority of $l$ is an unmatched $\textit{put}$ operation $o$, and the priority of $o$ equals the priority of some matched $\textit{put}$ operations in $l$, then $\textit{last}(l) = R_{\textit{pr4}}$.

    \item[-] If the maximal priority of $l$ is an unmatched $\textit{put}$ operation $o$, and the priority of $o$ is larger than the priority of all matched $\textit{put}$ operations in $l$, then $\textit{last}(l) = R_{\textit{pr3}}$.

    \item[-] If the maximal priority of $l$ is a matched $\textit{put}$ operation $o$, the priority of $o$ equals the priority of some matched $\textit{put}$ operations in $l$, and the priority of $o$ is larger than the priority of all unmatched $\textit{put}$ operations in $l$, then $\textit{last}(l) = R_{\textit{pr2}}$.

    \item[-] If the maximal priority of $l$ is an matched $\textit{put}$ operation $o$, and the priority of $o$ is larger than the priority of all $\textit{put}$ operations in $l$, then $\textit{last}(l) = R_{\textit{pr1}}$.

    \item[-] Otherwise, $l = \epsilon$, and $\textit{last}(l) = R_{\textit{pr0}}$.
    \end{itemize}
\end{itemize}

%The following lemma states that the inductive rules for priority queues are sound and complete.

\section{Step-by-Step Linearizability of Priority Queues}
\label{sec:step-by-step linearizability of priority queues}

The following lemma states that $\textit{PQueue}$ is closed under projection. This lemma holds since the predicates appears in inductive rules of $\textit{PQueue}$ are closed under projection.

\begin{restatable}{lemma}{PQisClosedUnderProjection}
\label{lemma:PQ is closed under projection}

$\textit{PQueue}$ is closed under projection.

\end{restatable}

%When the context is clear, we shorten priority-value data-independence as data-independence.
For priority queue, the notion of step-by-step linearizability is modified as follows:

%\vspace{-6pt}
\begin{definition}[step-by-step linearizability for priority queue]\label{def:step-by-step linearizability of priority queue}
A data structure $S=R_1, \ldots, R_n$ is said be to step-by-step linearizable, if for any data-differentiated sequence $e$, if $e$ is linearizable w.r.t. $\textit{MR}_{\textit{pri}}$ ($1 \leq i \leq 4$) with witness $x$, we have:

$e \setminus x \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket \Rightarrow e \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$

\noindent if $e$ is linearizable w.r.t. $\textit{MR}_{\textit{pr5}}$ with witness $x$, we have:

$e \setminus x \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr5}} \rrbracket \Rightarrow e \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr5}} \rrbracket$
\end{definition}

Let $\textit{proj}(s)$ denote the projection of $s$ obtained by erasing all the operations of several items from $s$. Here we assume that each $\textit{rm}(\textit{empty})$ is differentiated from each other, and we can consider each $\textit{rm}(\textit{empty})$ as having a unique item.

\begin{restatable}{lemma}{PQasMultiInMRpriforSequence}
\label{lemma:PQ as multi in MRpri for sequence}
$s \in \textit{PQueue}$, if and only if, $\forall s'$, if $s' \in \textit{proj}(s)$, then $s' \in \textit{MR}_{\textit{pri}}$ for some $0 \leq i \leq 5$.
\end{restatable}

\begin {proof}

The \textit{only if} direction is a direct consequence of Lemma \ref{lemma:PQ is closed under projection}.

To prove the $\textit{if}$ direction. The $\textit{if}$ direction holds trivially when $s = \epsilon$. When $s \neq \epsilon$. since $s \in \textit{proj}(s)$, we generate sequence $s_1$ from $s$ as follows:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $s$ has $\textit{rm}(\textit{empty})$ operations: $s_1$ is generated from $s$ by erasing some $\textit{rm}(\textit{empty})$ operation.

\item[-] If the maximal priority in $s$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in $S_1$ and unmatched $\textit{put}$ operations of some items in $S_2$: $s_1$ is generated from $s$ by erasing the unmatched $\textit{put}$ operation of items last putted in $S_2$.

\item[-] If the maximal priority in $s$ is unmatched $\textit{put}$ operations of some items in $S_2$ and $\vert S_2 \vert =1$: $s_1$ is generated from $s$ by erasing this unmatched $\textit{put}$ operation.

\item[-] If the maximal priority in $s$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in set $S$ and $\vert S \vert \geq 2$: $s_1$ is generated from $s$ by erasing matched $\textit{put}$ and $\textit{rm}$ operations of the item which is last putted in $S$.

\item[-] If the maximal priority in $s$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in set $S$ and $\vert S \vert = 1$: $s_1$ is generated from $s$ by erasing matched $\textit{put}$ and $\textit{rm}$ operations of the item in $S$.
\end{itemize}

Similarly, we obtain $s_2$ from $s_1$, obtain $s_3$ from $s_2$, $\ldots$, until we obtain some $s_m = \epsilon$. It is obvious that $s_m \in \textit{PQueue}$, and by the five rules of $\textit{PQueue}$, it is easy to see that $s_{\textit{m-1}},\ldots,s_1,s \in \textit{PQueue}$. \qed
\end {proof}

\begin{restatable}{lemma}{PQasMultiInMRpriforHistory}
\label{lemma:PQ as multi in MRpri for history}
Given a data-differentiated history $h$, assume that $\textit{PQueue}$ is step-by-step linearizability. $h \sqsubseteq \textit{PQueue}$, if and only if, $\forall h'$, if $h' \in \textit{proj}(h)$, then $h' \sqsubseteq \textit{MR}_{\textit{pri}}$, where $i = \textit{last}(h')$.
\end{restatable}

\begin {proof}

The \textit{only if} direction is a direct consequence of Lemma \ref{lemma:PQ is closed under projection}.

To prove the $\textit{if}$ direction. Since $h \sqsubseteq \textit{PQueue}$, we can see that $h \sqsubseteq l \in \textit{PQueue}$. Each projection $h'$ of $h$ can be linearizable to some projection $l'$ of $l$. By Lemma \ref{lemma:PQ as multi in MRpri for sequence}, $l' \in \textit{MR}_{\textit{prj}}$ for some $0 \leq j \leq 5$. Since $\textit{last}$ of a sequence is uniquely determined, we can see that $l' \in \textit{MR}_{\textit{pri}}$, where $i = \textit{last}(h')$.

The $\textit{if}$ direction holds trivially when $h = \epsilon$. When $h \neq \epsilon$. since $h \in \textit{proj}(h)$, we generate sequence $h_1$ from $h$ as follows:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $h$ has $\textit{rm}(\textit{empty})$ operations: Then $\textit{last}(h) = R_{\textit{pr5}}$, and $h \sqsubseteq \textit{MR}_{\textit{pr5}}$. $h_1$ is generated from $h$ by erasing some $\textit{rm}(\textit{empty})$ operation.

\item[-] If the maximal priority in $h$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in $S_1$ and unmatched $\textit{put}$ operations of some items in $S_2$: Then $\textit{last}(h) = R_{\textit{pr4}}$, and $h \sqsubseteq \textit{MR}_{\textit{pr4}}$. $h_1$ is generated from $h$ by erasing a unmatched $\textit{put}(x,\_)$, such that $\textit{put}(x,\_)$ does not happen before $\textit{put}(y,\_)$ for each $y \in S_2$.

\item[-] If the maximal priority in $s$ is unmatched $\textit{put}$ operations of some items in $S_2$ and $\vert S_2 \vert =1$: Then $\textit{last}(h) = R_{\textit{pr3}}$, and $h \sqsubseteq \textit{MR}_{\textit{pr3}}$. $h_1$ is generated from $h$ by erasing this unmatched $\textit{put}$ operation.

\item[-] If the maximal priority in $s$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in set $S$ and $\vert S \vert \geq 2$: Then $\textit{last}(h) = R_{\textit{pr2}}$, and $h \sqsubseteq \textit{MR}_{\textit{pr2}}$. $h_1$ is generated from $h$ by erasing $\textit{put}(x,\_)$ and $\textit{rm}(x)$, such that for each $y \in S$, $\textit{put}(x,\_)$ does not happen before $\textit{put}(y,\_)$, and $\textit{rm}(x)$ does not happen before $\textit{rm}(y)$.

\item[-] If the maximal priority in $s$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in set $S$ and $\vert S \vert = 1$: Then $\textit{last}(h) = R_{\textit{pr1}}$, and $h \sqsubseteq \textit{MR}_{\textit{pr1}}$. $h_1$ is generated from $h$ by erasing matched $\textit{put}$ and $\textit{rm}$ operations of the item in $S$.
\end{itemize}

Similarly, we obtain $h_2$ from $h_1$, obtain $h_3$ from $h_2$, $\ldots$, until we obtain some $h_m = \epsilon$. Assume that for each $i$, when generating $h_i$ from $h_{\textit{i-1}}$, we erase operations of item $d_i$. It is obvious that $h_m = h_{\textit{m-1}} \setminus d_m \sqsubseteq \textit{PQueue}$, and since $h_{\textit{m-1}} \sqsubseteq \textit{MR}_{\textit{prj}_{\textit{m-1}}}$, where $\textit{last}(h_{\textit{m-1}}) = R_{\textit{prj}_{\textit{m-1}}}$, we can see that $h_{\textit{m-1}} \sqsubseteq \textit{PQueue}$. Similarly, we can prove that $h_{\textit{m-1}},\ldots,h_1,h \sqsubseteq \textit{PQueue}$. This completes the proof of the $\textit{if}$ direction. \qed
\end {proof}

To prove that $\textit{PQueue}$ is step-by-step linearizability, we use five lemmas for five rules of $\textit{PQueue}$.

\begin{restatable}{lemma}{Rpr1isStepByStepLinearizability}
\label{lemma:Rpr1 is step-by-step linearizability}

$R_{\textit{pr1}}$ is step-by-step linearizability.

\end{restatable}

\begin {proof}

Assume that $l$ matches $R_{\textit{pr1}}$ with witness $x$. Assume that  Let $h$ be a data-differentiated history, and $l$ be an operation sequence such that $h \sqsubseteq l$. Then, let $h'=h \setminus x$ and assume that $h' \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$. Let $l'$ be a sequence in $\llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$ such that $h' \sqsubseteq l'$.

According to $R_{\textit{pr1}}$, there exist sequences $u$, $v$, and $w$, such that $l=u \cdot \textit{put}(x,\textit{pri}_x) \cdot v \cdot \textit{rm}(x) \cdot w$, where the $\textit{put}$ operations and $\textit{rm}$ operations in $u \cdot v$ are matched, and $\textit{pri}_x$ is larger than any priority in $u \cdot v \cdot w$. Let $D_L$ be the set of items in $u \cdot v$, let $D_R$ be the set of items in $w$. Since $u \cdot v$ contain only matched $\textit{put}$ and $\textit{rm}$ operations, we can see that $D_L \cap D_R = \emptyset$. Let $l'_L = l' \vert_{D_L}$ and $l'_R = l' \vert_{D_R}$. Let $S_1$ be the set of all operations of $h$ that happen before $\textit{put}(x,\textit{pri}_x)$, let $l'_{L1}$ be the shortest prefix of $l'_L$ that contains all operations of $S_1$, and let $l'_{L2}$ be the sequence such that $l'_L = l'_{L1} \cdot l'_{L2}$. It is obvious that if $\textit{put}(x,\textit{pri}_x)$ happens before an operation $o$ in $h$, then $o$ is in $l'_{L2}$. Let sequence $l'' = l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \cdot L'_R$.

Since priority queue is closed under projection (Lemma \ref{lemma:PQ is closed under projection}) and $u \cdot v$ contain matched $\textit{put}$ and $\textit{rm}$ operations, we know that $l'_{L1} \cdot l'_{L2} \in \textit{PQueue}$ and $l'_R \in \textit{PQueue}$. Since $\textit{pri}_x$ is larger than any priority in $h$, it is easy to see that $l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \in \textit{PQueue}$. Since the priority queue is empty after executing $l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \in \textit{PQueue}$, we know that $l'' = l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \cdot L'_R \in \textit{PQueue}$.

It remains to prove that $h \sqsubseteq l''$. To prove this, we define a graph $G$ whose nodes are the operations of $h$ and there is an edge from operation $o_1$ to $o_2$, if one of the following case holds

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $o_1$ happens-before $o_2$ in h,

\item[-] the method event corresponding to $o_1$ in $l''$ is before the one corresponding to $o_2$.
\end{itemize}

Assume there is a cycle in $G$. According the the property of interval order and the fact that the order of $l''$ is total, we know that there must exists $o_1$ and $o_2$, such that $o_1$ happens-before $o_2$ in $h$, but the corresponding method events are in the opposite order in $l''$. Then, we can see

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $o_1,o_2 \in l'_{L1}$, or $o_2 \in l'_{L1} \wedge o_1 \in l'_{L2}$, or $o_1,o_2 \in l'_{L2}$, or $o_1,o_2 \in L'_R$, then $l'$ contradicts with happen before relation of $h$.

\item[-] If $o_1 = \textit{put}(x,\textit{pri}_x) \wedge o_2 \in l'_{L1}$, then since $\textit{put}(x,\textit{pri}_x)$ happens before $o_2$, we know that $o_2 \in l'_{L2}$, which contradicts that $o_2 \in l'_{L1}$.

\item[-] If $o_1 = \textit{rm}(x) \wedge o_2 \in l'_{L1}$, or $o_1 \in L'_R \wedge o_2 \in l'_{L1}$, or $o_1 = \textit{rm}(x) \wedge o_2 = \textit{put}(x,\textit{pri}_x)$, or $o_1 \in L'_R \wedge o_2 = \textit{put}(x,\textit{pri}_x)$, or $o_1 = \textit{rm}(x) \wedge o_2 \in l'_{L2}$, or $o_1 \in L'_R \wedge o_2 \in l'_{L2}$, or $o_1 \in L'_R \wedge o_2 = \textit{rm}(x)$, then $l$ contradicts with happen before relation of $h$.

\item[-] If $o_1 \in l'_{L2} \wedge o_2 = \textit{put}(x,\textit{pri}_x)$, then since $o_1$ happens before $\textit{put}(x,\textit{pri}_x)$, we know that $o_1 \in S_1$ and then $o_1 \in l'_{L1}$, which contradicts that $o_1 \in l'_{L2}$.
\end{itemize}

Therefore, we know that $G$ is acyclic, and then we know that $h \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$. \qed

\end {proof}

\begin{restatable}{lemma}{Rpr2isStepByStepLinearizability}
\label{lemma:Rpr2 is step-by-step linearizability}

$R_{\textit{pr2}}$ is step-by-step linearizability.

\end{restatable}

\begin {proof}

Assume that $l$ matches $R_{\textit{pr2}}$ with witness $x$. Let $h$ be a data-differentiated history, and $l$ be an operation sequence such that $h \sqsubseteq l$. Let $h'=h \setminus x$ and assume that $h' \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$. Let $l'$ be a sequence in $\llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$ such that $h' \sqsubseteq l'$.

Since $h' \sqsubseteq l'$, there exists a sequence $h_{l'}$ of call, return and operations, such that (1) the projection of $h_{l'}$ into call and return action is $h$, (2) the projection of $h_{l'}$ into operations is $l'$, and (3) for each matched pair of call action $c$ and return action $r$ of $h$, there exists an operation $o$, such that $a$ before $o$, $o$ before $r$, and $o$ is the operation of $c$ and $r$. Intuitively, $h_{l'}$ is obtained from $h$ by adding linearization points.

According to $R_{\textit{pr2}}$, there exist sequences $u$, $v$, and $w$, such that $l=u \cdot \textit{put}(x,\textit{pri}_x) \cdot v \cdot \textit{rm}(x) \cdot w$, where (1) the $\textit{put}$ operations and $\textit{rm}$ operations in $u \cdot v$ are matched, (2) $\textit{pri}_x$ is the maximal priority in $u \cdot v \cdot w$, (3) there exists matched $\textit{put}$ and $\textit{rm}$ operations in $u \cdot v \cdot w$ that has priority $\textit{pri}_x$, (4) $\textit{pri}_x$ is larger than priorities of any unmatched put in $u \cdot v \cdot w$, and (5) all $\textit{put}$ operation with priority $\textit{pri}_x$ is in $u$. Let $D_L$ be the set of items in $u \cdot v$, let $D_R$ be the set of items in $w$. Since $u \cdot v$ contain only matched $\textit{put}$ and $\textit{rm}$ operations, we can see that $D_L \cap D_R = \emptyset$. Let $l'_L = l' \vert_{D_L}$ and $l'_R = l' \vert_{D_R}$.

Let $l'_{\textit{L0}}$ be the shortest prefix of $l'_L$, such that every operation in $l'_{\textit{L0}}$ is before the return action of $\textit{put}(x,\textit{pri}_x)$ in $h_{l'}$. Let $\textit{OP}_x = \{ \textit{put}(d,\textit{pri}_x) \vert \textit{put}(d,\textit{pri}_x)$ is an operation of $h\}$. Let $l'_{\textit{L1}} = l'_{\textit{L0}} \cdot l_t$, where $l_t$ is the projection of $l'$ into operations of $\textit{OP}_x$ that are not in $l'_{\textit{L0}}$. Let $\textit{OP}'_x$ be the set of operations in $l_t$. Let $l'_{\textit{L2}}$ be the projection of $l'_L$ into operations of $l'_L$ that are not in $l'_{\textit{L1}}$. Let sequence $l'' = l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \cdot L'_R$.

As in Lemma \ref{lemma:Rpr1 is step-by-step linearizability}, we know that $l'_L \in \textit{PQueue}$ and $l'_R \in \textit{PQueue}$. It is obvious that $l'_{\textit{L0}} \in \textit{PQueue}$, and then $l'_{\textit{L1}} = l'_{\textit{L0}} \cdot l_t \in \textit{PQueue}$. Since there is no $\textit{rm}(\textit{Empty})$ in $h$ and $\textit{pri}_x$ is the maximal priority of $h$, $l'_{L1} \cdot l'_{L2} \in \textit{PQueue}$. Since all $\textit{put}$ operations with priority $\textit{pri}$ is in $l'_{L1}$, we know that $l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \in \textit{PQueue}$. As in Lemma \ref{lemma:Rpr1 is step-by-step linearizability}, we know that $l'' \in \textit{PQueue}$.

To prove $h \sqsubseteq l''$, we define graph $G$ as in Lemma \ref{lemma:Rpr1 is step-by-step linearizability}. Assume that there is a cycle in $G$, then there must exists $o_1$ and $o_2$, such that $o_1$ happens-before $o_2$ in $h$, but the corresponding method events are in the opposite order in $l''$. Then, we can see

\begin{itemize}
\setlength{\itemsep}{0.5pt}

\item[-] $o_1 \in L'_R \wedge (o_2 \in l'_{L1} \vee o_2 = \textit{put}(x,\textit{pri}_x) \vee o_2 \in l'_{L2} \vee o_2 = \textit{rm}(x))$, or $o_1 = \textit{rm}(x) \wedge (o_2 \in l'_{L1} \vee o_2 = \textit{put}(x,\textit{pri}_x) \vee o_2 \in l'_{L2})$: Then $l$ contradicts with happen before relation of $h$.

\item[-] $o_1, o_2 \in l'_R$ or $o_1, o_2 \in l'_{L2}$: Then $l'$ contradicts with happen before relation of $h$.

\item[-] $o_1, o_2 \in l'_{L1}$: If $(o_1, o_2 \in l'_{L0}) \vee (o_1,o_2 \in \textit{OP}'_x)$, then $l'$ contradicts with happen before relation of $h$.

    Else, it is the case that $o_1 \in \textit{OP}'_x \wedge o_2 \in l'_{L0}$. We can see that $o_1$ interleaves with $\textit{put}(x,\textit{pri}_x)$ and the linearization point of $o_1$ is after the return action of $\textit{put}(x,\textit{pri}_x)$. Since $o_1 <_{\textit{hb}} o_2$, it is easy to see that $\textit{put}(x,\textit{pri}_x) <_{\textit{hb}} o_2$, and then $o_2 \notin l'_{L0}$, which contradicts that $o_2 \in l'_{L0}$.

\item[-] $o_1 = \textit{put}(x,\textit{pri}_x) \wedge o_2 \in l'_{L1}$: Since $o_1 <_{\textit{hb}} o_2$, it is obvious that $o_2 \notin l'_{L1}$, contradicts that $o_2 \in l'_{L1}$.

\item[-] $o_1 \in l'_{L2} \wedge o_2 \in l'_{L1}$: If $o_2 \in l'_{L0}$, then $l'$ contradicts with happen before relation of $h$.

    Else, it is the case that $o_2 \in \textit{OP}'_x$. $o_2$ interleaves with $\textit{put}(x,\textit{pri}_x)$ and the linearization point of $o_2$ is after the return action of $\textit{put}(x,\textit{pri}_x)$. Since $o_1 <_{\textit{hb}} o_2$, the return action of $o_1$ is before the return action of $\textit{put}(x,\textit{pri}_x)$. Then it is easy to see that $o_1 \in l'_{L0}$, contradicts that $o_1 \in l'_{L2}$.

\item[-] $o_1 \in l'_{L2} \wedge o_2 = \textit{put}(x,\textit{pri}_x)$: Then $o_1 \in l'_{L0}$, contradicts that $o_1 \in l'_{L2}$.
\end{itemize}

Therefore, we know that $G$ is acyclic, and then we know that $h \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$. \qed
\end {proof}

\begin{restatable}{lemma}{Rpr3isStepByStepLinearizability}
\label{lemma:Rpr3 is step-by-step linearizability}

$R_{\textit{pr3}}$ is step-by-step linearizability.

\end{restatable}

\begin {proof}

>>>>

\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{color}
\usepackage{amsfonts,amssymb}
%\usepackage{txfonts}
%\usepackage{pxfonts}
%\usepackage{mathabx}
%\usepackage{makeidx}  % allows for indexgeneration
\usepackage{verbatim}
\usepackage{url}
%\usepackage{hyperref}
\usepackage{times}
\usepackage{ulem}    %(\emph{....} in the texfile) is underlined instead of italic
\normalem
\usepackage{enumerate}
%\usepackage[square, comma, sort&compress, numbers]{natbib}

@@

\author {Ahmed Bouajjani \and Michael Emmi \and Constantin Enea \and Chao Wang}
\institute{Institut de Recherche en Informatique Fondamentale, \\Univ. Paris Diderot (Paris 7)
}

@@

\input{abstract}

\input{introduction}

\input{preliminaries}

\input{data-independence-PQ}

\input{rules-PQ}
